Integrated Ocular Million Single Cell Meta-Atlas for Discovery and Development

### 2020-08-04 manuscript_00.Rmd
text update

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To curate a accessible and reliable ocular cell-type specific transcriptome dataset of over one million normal vertebrate ocular cell types with a matching high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We queried several public repositories to find over a dozen healthy, non-perturbed, ocular tissue single cell RNA-seq studies As a non-ocular reference, we also downloaded the pan-mouse Tabula Muris raw data. The XX ocular and XX Tabula Muris cells were sent into a Snakemake-based reproducible pipeline we wrote to quantify transcripts, remove cells with poor quality, normalize expression values, creates a batch-corrected low dimensional space, performs XX differential expression tests, and outputs all as a single SQLite database file: the Single Cell Eye in a Disk (scEiaD) dataset. XXX thousand published cell labels assignments were curated and a machine learning  projected onto the non-labelled cells. Furthermore, we wrote the web application PLAE (https://plae.nei.nih.gov) to display scEiaD. RESULTS: The new portal provides quick visualization of vertebrate eye-related transcriptomes published to date by gene/transcript, XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we showed XX. CONCLUSION: The PLAE v1.0 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues. Futhermore the XXX labelled cells across XX studies, X techologies, and 3 species provides a highly challenging benchmarking resource for single cell algorithm development."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To curate a accessible and reliable ocular cell-type specific transcriptome dataset of over one million normal vertebrate ocular cell types with a matching high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We queried several public repositories to find eighteen healthy, non-perturbed, ocular tissue single cell RNA-seq studies As a non-ocular reference, we also used the pan-mouse Tabula Muris resource. The XX ocular and XX Tabula Muris cells were sent into a Snakemake-based reproducible pipeline we wrote to quantify transcripts, remove cells with poor quality, normalize expression values, create a batch-corrected low dimensional space, performs XX differential expression tests, and outputs all as a single SQLite database file: the Single Cell Eye in a Disk (scEiaD) dataset. XXX thousand published cell labels assignments were curated and a machine learning  projected onto the non-labelled cells. Furthermore, we wrote the web application PLAE (https://plae.nei.nih.gov) to display scEiaD. RESULTS: The new portal provides quick visualization of vertebrate eye-related transcriptomes published to date by gene/transcript, XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we showed XX. CONCLUSION: The PLAE v1.0 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues. Futhermore the XXX labelled cells across XX studies, X techologies, and 3 species provides a highly challenging benchmarking resource for single cell algorithm development."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

# Introduction
## Many published ocular atlases

## Data not easily available for all

## Some have accompanying apps

## But apps are generally slow and have minimal exploration tooling

## Cross comparisons difficult

## If exploring a gene expression pattern, crucial to have reproducibility across disparate groups to enhance confidence that effect is biological

## Single cell datasets getting larger
Valentine Svensson figure?
MOCA atlas (mouse 2e6 cells?)

## Thus crucial for algorithms to be highly performant
Our multidimensional atlas of 1e6 cells will be a useful reference dataset in:
  - batch effect correction tools
  - clustering algorithms
  - cell type prediction / projection
  - trajectory inference

## What we did
We have built an integrated meta-atlas that:

  Has a batch corrected lower dimensional space (8 dimensions) from scVI

  Curated previously labelled cell types for >400,000 cells.

  Projected these onto the remaining unlabelled cells.

  Made 77 clusters
   - subclustered each to create XX total clusters

  Pre-calculated pseudobulk differential expression testing in 9 different testing schemes:

## Will do???

Trajectory

Velocity

# Methods

## Identify ocular scRNA studies
SRA
EBI
Pubmed
Found 18 (?) studies

## pan mouse as non-ocular reference
Tabula Muris

## Quantification

kallisto/bustools

## Batch normalization
Tested:
  - scVI
  - CCA
  - insct
  - combat
  - nothing
  - magic
  - scanorama
  - harmony
  - fastMNN

## Grid search to pick optimal parameters

Varied:
  - number of Hyper Variable Genes (HVG): 1000, 2000,5000,1000
  - dims: 4,6,8,10,20,30,50,100
  - normalization: scran, seurat standard, libSize, sqrt, counts (scVI only)
  - nneighbors (UMAP viz only)
  - knn: 5,7,10 (clustering)

benchmarked with:
  - LISI
  - ARI
  - Silhouette
  - PCA variance explained before/after correction

## Clustering
Louvain-jaccard from Seurat

##

# Results

## pseudobulk differential expression testing
  - celltype (published) against remaining
  - celltype (published) pairwise against celltype (published)
  - species specific within celltype
  - same for celltype (projected)
  - same for cluster

# Conclusion
